The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-08 rational.tex

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type $a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

Now consider the following algebraic manipulation (assuming $a \neq 0$):
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
\end{array}\]
As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
{{a <-> a}} isomorphism computes as follows:
\begin{itemize}
\item We start at line (0) with the value {{20}};
\item We proceed to line (1) with the value {{(20,())}};
\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
  $\alpha$ is essentially a logic variable;
\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
  forces the unification $\alpha=20$, i.e., the value becomes
  {{((20,1/20),20)}} which simplifies to {{((),20)}};
\item We reach line (5) with the value {{20}}.
\end{itemize}

\amr{the above is a boring trace; we need to convey why it is exciting but we
  are not agreeing on the interpretation.}

>>>>

start with int times bool divided by bool: it is like an int but it can only
be used as the second component of a pair whose first component is an int;
gives you the power to say here is 20 but you must use it to buy books

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~$a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

% Now consider the following algebraic manipulation (assuming $a \neq 0$):
% \[\begin{array}{rclll}
% && a & \mbox{initial~$a$} & (0) \\
% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
% \end{array}\]
% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
% {{a <-> a}} isomorphism commputes as follows:
% \begin{itemize}
% \item We start at line (0) with the value {{20}};
% \item We proceed to line (1) with the value {{(20,())}};
% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
%   $\alpha$ is essentially a logic variable;
% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
%   forces the unification $\alpha=20$, i.e., the value becomes
%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
% \item We reach line (5) with the value {{20}}.
% \end{itemize}
% \amr{the above is a boring trace; we need to convey why it is exciting but we
%   are not agreeing on the interpretation.}
